---
title: "Methods Proposal and Outline"
author: "Yiming Cao, Zichen Fan, Xinyin Miao, Teresa Sha"
date: "2025-03-25"
output:
  pdf_document:
    latex_engine: xelatex
---

## EDA recap and motivation

From the EDA we found strong seasonal and day-of-week effects on mental health EMS calls, a nonlinear temperature–outcome relationship, and only a weak PM2.5 signal in raw scatter plots. The data has 124 observations (EPA monitors sample every 3 days) and the outcome is a daily count (mental_health mean ~486, suicide mean ~5), so we need a count regression that can handle overdispersion without overfitting on a small n.

This motivates a quasi-Poisson time series GLM — it handles overdispersed counts, lets us flexibly adjust for time trends and temperature with splines, and only requires base R packages.

## Quasi-Poisson GLM — analysis plan (Yiming Cao)

### Model

$$
\log(\mu_t) = \beta_0 + \beta_1 \cdot \text{PM2.5}_t + \text{ns}(\text{time}, df) + \text{ns}(\text{tmean}_t, 3) + \beta_2 \cdot \text{is\_weekend}_t + \beta_3 \cdot \text{precip}_t
$$

where $Y_t \sim \text{Quasi-Poisson}$. Same structure for both outcomes (mental_health and suicide).

### Covariate selection

- **ns(time, 4)**: natural spline to capture the seasonal trend seen in EDA. ~1 df per 3 months.
- **ns(tmean, 3)**: nonlinear temperature adjustment. EDA showed a curved relationship that a linear term wouldn't capture, 3 df should be sufficient given n = 124.
- **is_weekend**: binary indicator. EDA showed a clear weekday/weekend split (~510 vs ~425). Using full factor(weekday) would cost 6 df which seems excessive for 124 obs.
- **precip**: linear term as a nuisance covariate — precipitation may affect both air quality and behavior.
- Not including wind speed or RH to avoid collinearity issues with the small sample size.

### R packages

`stats` (glm), `splines` (ns), `ggplot2` (figures).

### Sensitivity analyses

- Vary time trend spline df from 3 to 6 to check whether the PM2.5 estimate is sensitive to how much seasonal variation we absorb
- Exclude high-PM2.5 days (>95th percentile) that are likely wildfire smoke events, which represent a different exposure pattern
- Test lagged temperature (previous measurement day) instead of same-day

### Expected visualizations

- Rate ratio forest plot: PM2.5 effect (per 10 μg/m³) with 95% CI for both outcomes
- Sensitivity plot showing how the PM2.5 coefficient changes across different time trend df choices
- Residual diagnostics: deviance residuals vs fitted, PACF of residuals to check for remaining autocorrelation

## Methods section outline

- Quasi-Poisson GLM to estimate association between same-day PM2.5 and mental health EMS call counts in NYC, 2024
- Primary outcome: daily mental health crisis calls (EDP + ALTMEN); secondary: suicide-related calls (JUMPDN + JUMPUP)
- Exposure: city-wide mean PM2.5 from EPA FRM monitors
- Adjusted for: seasonal/long-term trend (natural spline, 4 df), temperature (natural spline, 3 df), weekend indicator, precipitation
- Results as rate ratio per 10 μg/m³ PM2.5 increase with 95% CIs
- Sensitivity: time trend df (3–6), excluded wildfire days, lagged temperature

## Team assignments

| Method | Person | Status |
|--------|--------|--------|
| Quasi-Poisson GLM | Yiming Cao | in progress |
| Mixed Model | Zichen Fan | — |
| Case-crossover | Teresa Sha | — |
| TBD | Xinyin Miao | — |



############################################################
## case_crossover_sensitivity.R----Teresa Sha
## Sensitivity analysis:
## time-stratified case-crossover using conditional Poisson
############################################################

## ---------------------------
## 0. packages
## ---------------------------
packages <- c("dplyr", "lubridate", "splines", "gnm", "broom")

to_install <- packages[!packages %in% installed.packages()[, "Package"]]
if (length(to_install) > 0) install.packages(to_install)

library(dplyr)
library(lubridate)
library(splines)
library(gnm)
library(broom)

```

```{r}
## ---------------------------
## 1. read data
## ---------------------------
## Choose ONE path that matches your own folder structure.

## Option A: if your script is inside the project folder
## and analysis_2024.csv is in data/
file_path <- "data/analysis_2024.csv"

## Option B: if analysis_2024.csv is in the same folder as this script
## file_path <- "analysis_2024.csv"

## Option C: if you want to use full local path on your computer
## file_path <- "/Users/yourname/Desktop/aa_project/data/analysis_2024.csv"

file_path <- file.choose()
df <- read.csv(file_path)

```

```{r}
## ---------------------------
## 2. basic cleaning
## ---------------------------
df <- df %>%
  mutate(
    date = as.Date(date),
    year = year(date),
    month_num = month(date),
    dow_num = wday(date, week_start = 1),  # Monday = 1
    dow_name = factor(
      dow_num,
      levels = 1:7,
      labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  ) %>%
  arrange(date)

## Quick check
cat("Rows:", nrow(df), "\n")
cat("Date range:", as.character(min(df$date)), "to", as.character(max(df$date)), "\n")
print(names(df))
```

```{r}
## ---------------------------
## 3. create case-crossover strata
## ---------------------------
## Time-stratified control selection:
## same year + same month + same day of week
## This matches the usual case-crossover setup.

df <- df %>%
  mutate(
    stratum = interaction(year, month_num, dow_name, drop = TRUE)
  )

## Keep only strata with at least 2 observed days
## because strata with only 1 row do not contribute within-stratum comparison
df_cc <- df %>%
  group_by(stratum) %>%
  mutate(n_in_stratum = n()) %>%
  ungroup() %>%
  filter(n_in_stratum >= 2)

cat("Rows after dropping singleton strata:", nrow(df_cc), "\n")
cat("Number of strata kept:", n_distinct(df_cc$stratum), "\n")
```

```{r}
## ---------------------------
## 4. create lag variables
## ---------------------------
## Since PM2.5 is measured only on selected days,
## these lags are based on previous observed row in the merged dataset.
## For this sensitivity analysis, lag 0 and lag 1 are enough.

df_cc <- df_cc %>%
  arrange(date) %>%
  mutate(
    pm25_lag1 = lag(pm25, 1),
    pm25_lag2 = lag(pm25, 2),
    tmean_lag1 = lag(tmean, 1),
    rh_max_lag1 = lag(rh_max, 1)
  )

## Datasets for lag models
df_cc_lag1 <- df_cc %>%
  filter(!is.na(pm25_lag1), !is.na(tmean_lag1), !is.na(rh_max_lag1))

df_cc_lag2 <- df_cc %>%
  filter(!is.na(pm25_lag1), !is.na(pm25_lag2),
         !is.na(tmean_lag1), !is.na(rh_max_lag1))
```

```{r}
## ---------------------------
## 5. helper function for IRR output
## ---------------------------
get_irr <- function(model, exposure_name = "pm25", increment = 1) {
  beta <- coef(model)[exposure_name]
  se   <- sqrt(vcov(model)[exposure_name, exposure_name])

  est  <- exp(beta * increment)
  low  <- exp((beta - 1.96 * se) * increment)
  high <- exp((beta + 1.96 * se) * increment)

  out <- data.frame(
    term = exposure_name,
    increment = increment,
    IRR = est,
    CI_low = low,
    CI_high = high
  )
  return(out)
}

```

```{r}
## ---------------------------
## 6. main conditional Poisson models
## ---------------------------
## Outcome 1: mental_health
## Model 1: lag 0 only
m_mh_lag0 <- gnm(
  mental_health ~ pm25 + ns(tmean, df = 3) + ns(rh_max, df = 3),
  eliminate = stratum,
  family = quasipoisson(link = "log"),
  data = df_cc
)
```

```{r}
## Model 2: lag 0 + lag 1
m_mh_lag01 <- gnm(
  mental_health ~ pm25 + pm25_lag1 + ns(tmean, df = 3) + ns(rh_max, df = 3),
  eliminate = stratum,
  family = quasipoisson(link = "log"),
  data = df_cc_lag1
)
```

```{r}

## Model 3: interaction with heat_day
m_mh_interaction <- gnm(
  mental_health ~ pm25 * heat_day + ns(tmean, df = 3) + ns(rh_max, df = 3),
  eliminate = stratum,
  family = quasipoisson(link = "log"),
  data = df_cc
)
```

```{r}
## Outcome 2: suicide
## Model 4: lag 0 only
m_su_lag0 <- gnm(
  suicide ~ pm25 + ns(tmean, df = 3) + ns(rh_max, df = 3),
  eliminate = stratum,
  family = quasipoisson(link = "log"),
  data = df_cc
)
```

```{r}
## Model 5: lag 0 + lag 1
m_su_lag01 <- gnm(
  suicide ~ pm25 + pm25_lag1 + ns(tmean, df = 3) + ns(rh_max, df = 3),
  eliminate = stratum,
  family = quasipoisson(link = "log"),
  data = df_cc_lag1
)
```

```{r}
## ---------------------------
## 7. optional: two-pollutant-style lag sensitivity
## ---------------------------
## This one is optional. Use only if you want one extra robustness check.
m_mh_lag012 <- gnm(
  mental_health ~ pm25 + pm25_lag1 + pm25_lag2 +
    ns(tmean, df = 3) + ns(rh_max, df = 3),
  eliminate = stratum,
  family = quasipoisson(link = "log"),
  data = df_cc_lag2
)
```

```{r}
## ---------------------------
## 8. model summaries
## ---------------------------
cat("\n==============================\n")
cat("Mental health model: lag 0\n")
cat("==============================\n")
print(summary(m_mh_lag0))

cat("\n==============================\n")
cat("Mental health model: lag 0 + lag 1\n")
cat("==============================\n")
print(summary(m_mh_lag01))

cat("\n==============================\n")
cat("Mental health model: PM2.5 * heat_day\n")
cat("==============================\n")
print(summary(m_mh_interaction))

cat("\n==============================\n")
cat("Suicide model: lag 0\n")
cat("==============================\n")
print(summary(m_su_lag0))

cat("\n==============================\n")
cat("Suicide model: lag 0 + lag 1\n")
cat("==============================\n")
print(summary(m_su_lag01))

cat("\n==============================\n")
cat("Mental health model: lag 0 + lag 1 + lag 2\n")
cat("==============================\n")
print(summary(m_mh_lag012))
```

```{r}
## ---------------------------
## 9. IRR tables
## ---------------------------
irr_results <- bind_rows(
  cbind(model = "mental_health_lag0_1ug", get_irr(m_mh_lag0, "pm25", 1)),
  cbind(model = "mental_health_lag0_5ug", get_irr(m_mh_lag0, "pm25", 5)),
  cbind(model = "mental_health_lag01_pm25_1ug", get_irr(m_mh_lag01, "pm25", 1)),
  cbind(model = "mental_health_lag01_pm25lag1_1ug", get_irr(m_mh_lag01, "pm25_lag1", 1)),
  cbind(model = "suicide_lag0_1ug", get_irr(m_su_lag0, "pm25", 1)),
  cbind(model = "suicide_lag0_5ug", get_irr(m_su_lag0, "pm25", 5)),
  cbind(model = "suicide_lag01_pm25_1ug", get_irr(m_su_lag01, "pm25", 1)),
  cbind(model = "suicide_lag01_pm25lag1_1ug", get_irr(m_su_lag01, "pm25_lag1", 1))
)

cat("\n==============================\n")
cat("IRR results\n")
cat("==============================\n")
print(irr_results)

```

```{r}

## ---------------------------
## 10. save outputs
## ---------------------------
if (!dir.exists("results")) dir.create("results")

write.csv(
  irr_results,
  "results/case_crossover_irr_results.csv",
  row.names = FALSE
)

saveRDS(m_mh_lag0, "results/m_mh_lag0.rds")
saveRDS(m_mh_lag01, "results/m_mh_lag01.rds")
saveRDS(m_mh_interaction, "results/m_mh_interaction.rds")
saveRDS(m_su_lag0, "results/m_su_lag0.rds")
saveRDS(m_su_lag01, "results/m_su_lag01.rds")
saveRDS(m_mh_lag012, "results/m_mh_lag012.rds")

cat("\nSaved model outputs to /results folder.\n")
```

```{r}
## ---------------------------
## 11. optional compact coefficient table
## ---------------------------
coef_table <- bind_rows(
  tidy(m_mh_lag0) %>% mutate(model = "m_mh_lag0"),
  tidy(m_mh_lag01) %>% mutate(model = "m_mh_lag01"),
  tidy(m_mh_interaction) %>% mutate(model = "m_mh_interaction"),
  tidy(m_su_lag0) %>% mutate(model = "m_su_lag0"),
  tidy(m_su_lag01) %>% mutate(model = "m_su_lag01"),
  tidy(m_mh_lag012) %>% mutate(model = "m_mh_lag012")
)

write.csv(
  coef_table,
  "results/case_crossover_coef_table.csv",
  row.names = FALSE
)

cat("Saved coefficient table.\n")
```

