---
title: "Preliminary results"
author: "Yiming Cao, Zichen Fan, Xinyin Miao, Teresa Sha"
output:
  pdf_document:
    latex_engine: xelatex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


# Nonlinear Exposure--Response Analysis (Xinyin Miao)

## Load packages

```{r}
library(dplyr)
library(ggplot2)
library(splines)
library(lubridate)
library(purrr)
library(tidyr)
```

## Read and prepare data

```{r}
df <- read.csv("data/analysis_2024.csv") |>
  mutate(
    date = as.Date(date),
    time = seq_len(n()),
    month_name = factor(month(date, label = TRUE, abbr = TRUE),
                        levels = month.abb),
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5)  ~ "Spring",
      month(date) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                         ~ "Fall"
    ) |> factor(levels = c("Winter", "Spring", "Summer", "Fall")),
    weekday_name = factor(
      weekdays(date),
      levels = c("Monday", "Tuesday", "Wednesday",
                 "Thursday", "Friday", "Saturday", "Sunday")
    ),
    is_weekend = as.numeric(is_weekend)
  )

str(df)
summary(df[, c("pm25", "mental_health", "suicide", "tmean", "precip", "is_weekend")])
```

## Model specification

Primary nonlinear model:

$$
\log(\mu_t) = \beta_0 + \operatorname{ns}(\mathrm{PM}_{2.5,t}, 3)
+ \operatorname{ns}(\mathrm{time}_t, 4)
+ \operatorname{ns}(\mathrm{tmean}_t, 3)
+ \beta_1 \cdot \mathrm{is\_weekend}_t
+ \beta_2 \cdot \mathrm{precip}_t
$$

where

$$
Y_t \sim \text{Quasi-Poisson}, \qquad \mu_t = E(Y_t).
$$

For comparison, the linear specification is:

$$
\log(\mu_t) = \beta_0 + \beta_1 \cdot \mathrm{PM}_{2.5,t}
+ \operatorname{ns}(\mathrm{time}_t, 4)
+ \operatorname{ns}(\mathrm{tmean}_t, 3)
+ \beta_2 \cdot \mathrm{is\_weekend}_t
+ \beta_3 \cdot \mathrm{precip}_t
$$

## Fit linear and nonlinear models

### Mental health outcome

```{r}
fit_mh_linear <- glm(
  mental_health ~ pm25 + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df
)

fit_mh_nonlinear <- glm(
  mental_health ~ ns(pm25, 3) + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df
)

summary(fit_mh_linear)
summary(fit_mh_nonlinear)
```

### Suicide outcome

```{r}
fit_sui_linear <- glm(
  suicide ~ pm25 + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df
)

fit_sui_nonlinear <- glm(
  suicide ~ ns(pm25, 3) + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df
)

summary(fit_sui_linear)
summary(fit_sui_nonlinear)
```

## Model comparison

```{r}
model_compare <- data.frame(
  Model = c("MH linear", "MH nonlinear", "Suicide linear", "Suicide nonlinear"),
  Residual_Deviance = c(
    fit_mh_linear$deviance,
    fit_mh_nonlinear$deviance,
    fit_sui_linear$deviance,
    fit_sui_nonlinear$deviance
  ),
  DF_Residual = c(
    fit_mh_linear$df.residual,
    fit_mh_nonlinear$df.residual,
    fit_sui_linear$df.residual,
    fit_sui_nonlinear$df.residual
  ),
  Dispersion = c(
    summary(fit_mh_linear)$dispersion,
    summary(fit_mh_nonlinear)$dispersion,
    summary(fit_sui_linear)$dispersion,
    summary(fit_sui_nonlinear)$dispersion
  )
)

model_compare
```

```{r}
drop1(fit_mh_nonlinear, test = "F")
drop1(fit_sui_nonlinear, test = "F")
```

## Rate ratio per 10 \mu g/m^3 for linear model

```{r}
get_rr10 <- function(model, var = "pm25") {
  beta <- coef(model)[var]
  se   <- sqrt(vcov(model)[var, var])

  rr    <- exp(beta * 10)
  lower <- exp((beta - 1.96 * se) * 10)
  upper <- exp((beta + 1.96 * se) * 10)

  data.frame(
    term = var,
    beta = beta,
    RR_10 = rr,
    lower_95 = lower,
    upper_95 = upper
  )
}

rr_mh_linear  <- get_rr10(fit_mh_linear, "pm25")
rr_sui_linear <- get_rr10(fit_sui_linear, "pm25")

rr_mh_linear
rr_sui_linear
```

## Create exposure--response curve function

```{r}
make_er_curve <- function(model, data, outcome_name,
                          pm_var = "pm25",
                          ref_value = median(data[[pm_var]], na.rm = TRUE),
                          n_points = 200) {

  pm_seq <- seq(min(data[[pm_var]], na.rm = TRUE),
                max(data[[pm_var]], na.rm = TRUE),
                length.out = n_points)

  newdata <- data.frame(
    pm25 = pm_seq,
    time = median(data$time, na.rm = TRUE),
    tmean = median(data$tmean, na.rm = TRUE),
    is_weekend = 0,
    precip = median(data$precip, na.rm = TRUE)
  )

  refdata <- newdata[1, ]
  refdata$pm25 <- ref_value

  X_new <- model.matrix(delete.response(terms(model)), newdata)
  X_ref <- model.matrix(delete.response(terms(model)), refdata)

  beta <- coef(model)
  V    <- vcov(model)

  diff_mat <- X_new - matrix(X_ref, nrow = nrow(X_new), ncol = ncol(X_new), byrow = TRUE)

  eta_diff <- as.vector(diff_mat %*% beta)
  se_diff  <- sqrt(diag(diff_mat %*% V %*% t(diff_mat)))

  data.frame(
    pm25 = pm_seq,
    RR = exp(eta_diff),
    lower = exp(eta_diff - 1.96 * se_diff),
    upper = exp(eta_diff + 1.96 * se_diff),
    outcome = outcome_name,
    ref_value = ref_value
  )
}

make_linear_curve <- function(model, data, outcome_name,
                              pm_var = "pm25",
                              ref_value = median(data[[pm_var]], na.rm = TRUE),
                              n_points = 200) {

  pm_seq <- seq(min(data[[pm_var]], na.rm = TRUE),
                max(data[[pm_var]], na.rm = TRUE),
                length.out = n_points)

  beta <- coef(model)["pm25"]
  se   <- sqrt(vcov(model)["pm25", "pm25"])

  eta_diff <- beta * (pm_seq - ref_value)
  se_diff  <- abs(pm_seq - ref_value) * se

  data.frame(
    pm25 = pm_seq,
    RR = exp(eta_diff),
    lower = exp(eta_diff - 1.96 * se_diff),
    upper = exp(eta_diff + 1.96 * se_diff),
    outcome = outcome_name,
    model_type = "Linear"
  )
}
```

## Generate exposure--response curves

```{r}
curve_mh  <- make_er_curve(fit_mh_nonlinear, df, "Mental health")
curve_sui <- make_er_curve(fit_sui_nonlinear, df, "Suicide")
```

## Plot nonlinear exposure--response curves

```{r fig.width=8, fig.height=5}
ggplot(curve_mh, aes(x = pm25, y = RR)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2, color = "gray40") +
  labs(
    title = "PM2.5 exposure-response curve: Mental health EMS calls",
    subtitle = paste0("Reference PM2.5 = ", round(unique(curve_mh$ref_value), 2), " \u03bcg/m^3"),
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio"
  ) +
  theme_bw()
```

```{r fig.width=8, fig.height=5}
ggplot(curve_sui, aes(x = pm25, y = RR)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2, color = "gray40") +
  labs(
    title = "PM2.5 exposure-response curve: Suicide-related EMS calls",
    subtitle = paste0("Reference PM2.5 = ", round(unique(curve_sui$ref_value), 2), " \u03bcg/m^3"),
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio"
  ) +
  theme_bw()
```

## Compare linear vs nonlinear PM2.5 specification

```{r}
curve_mh_linear  <- make_linear_curve(fit_mh_linear, df, "Mental health")
curve_sui_linear <- make_linear_curve(fit_sui_linear, df, "Suicide")

curve_mh_nl_plot  <- curve_mh  |> mutate(model_type = "Nonlinear")
curve_sui_nl_plot <- curve_sui |> mutate(model_type = "Nonlinear")

curve_compare_mh  <- bind_rows(curve_mh_linear, curve_mh_nl_plot)
curve_compare_sui <- bind_rows(curve_sui_linear, curve_sui_nl_plot)
```

```{r fig.width=8, fig.height=5}
ggplot(curve_compare_mh, aes(x = pm25, y = RR, color = model_type)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2, color = "gray40") +
  labs(
    title = "Linear vs nonlinear PM2.5 relationship: Mental health calls",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio",
    color = "Model"
  ) +
  theme_bw()
```

```{r fig.width=8, fig.height=5}
ggplot(curve_compare_sui, aes(x = pm25, y = RR, color = model_type)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2, color = "gray40") +
  labs(
    title = "Linear vs nonlinear PM2.5 relationship: Suicide-related calls",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio",
    color = "Model"
  ) +
  theme_bw()
```

## Distribution of PM2.5

```{r fig.width=7, fig.height=4.5}
ggplot(df, aes(x = pm25)) +
  geom_histogram(bins = 20, fill = "gray70", color = "white") +
  geom_vline(xintercept = median(df$pm25, na.rm = TRUE), linetype = 2, color = "red") +
  labs(
    title = "Distribution of PM2.5 in analysis sample",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Count of observation days"
  ) +
  theme_bw()
```

## Residual diagnostics

```{r fig.width=7, fig.height=4.5}
diag_mh <- data.frame(
  fitted = fitted(fit_mh_nonlinear),
  resid  = residuals(fit_mh_nonlinear, type = "deviance")
)

ggplot(diag_mh, aes(fitted, resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray40") +
  labs(
    title = "Deviance residuals vs fitted: Mental health nonlinear model",
    x = "Fitted values",
    y = "Deviance residuals"
  ) +
  theme_bw()
```

```{r fig.width=7, fig.height=4.5}
diag_sui <- data.frame(
  fitted = fitted(fit_sui_nonlinear),
  resid  = residuals(fit_sui_nonlinear, type = "deviance")
)

ggplot(diag_sui, aes(fitted, resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray40") +
  labs(
    title = "Deviance residuals vs fitted: Suicide nonlinear model",
    x = "Fitted values",
    y = "Deviance residuals"
  ) +
  theme_bw()
```

```{r}
pacf(
  residuals(fit_mh_nonlinear, type = "pearson"),
  main = "PACF of Pearson residuals: Mental health nonlinear model"
)

pacf(
  residuals(fit_sui_nonlinear, type = "pearson"),
  main = "PACF of Pearson residuals: Suicide nonlinear model"
)
```

## Sensitivity analysis 1: PM2.5 spline df = 4

```{r}
fit_mh_nonlinear_df4 <- glm(
  mental_health ~ ns(pm25, 4) + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df
)

fit_sui_nonlinear_df4 <- glm(
  suicide ~ ns(pm25, 4) + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df
)

curve_mh_df4  <- make_er_curve(fit_mh_nonlinear_df4, df, "Mental health") |> mutate(df_pm = "df = 4")
curve_sui_df4 <- make_er_curve(fit_sui_nonlinear_df4, df, "Suicide") |> mutate(df_pm = "df = 4")

curve_mh_df3  <- curve_mh  |> mutate(df_pm = "df = 3")
curve_sui_df3 <- curve_sui |> mutate(df_pm = "df = 3")

curve_mh_sens  <- bind_rows(curve_mh_df3, curve_mh_df4)
curve_sui_sens <- bind_rows(curve_sui_df3, curve_sui_df4)
```

```{r fig.width=8, fig.height=5}
ggplot(curve_mh_sens, aes(pm25, RR, color = df_pm)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to PM2.5 spline df: Mental health",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio",
    color = "PM2.5 spline"
  ) +
  theme_bw()
```

```{r fig.width=8, fig.height=5}
ggplot(curve_sui_sens, aes(pm25, RR, color = df_pm)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to PM2.5 spline df: Suicide",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio",
    color = "PM2.5 spline"
  ) +
  theme_bw()
```

## Sensitivity analysis 2: exclude high PM2.5 days

```{r}
pm95 <- quantile(df$pm25, 0.95, na.rm = TRUE)

df_no_high <- df |> filter(pm25 <= pm95)

fit_mh_no_high <- glm(
  mental_health ~ ns(pm25, 3) + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df_no_high
)

fit_sui_no_high <- glm(
  suicide ~ ns(pm25, 3) + ns(time, 4) + ns(tmean, 3) + is_weekend + precip,
  family = quasipoisson(link = "log"),
  data = df_no_high
)

curve_mh_no_high <- make_er_curve(fit_mh_no_high, df_no_high, "Mental health") |>
  mutate(sample = "Exclude >95th percentile")

curve_sui_no_high <- make_er_curve(fit_sui_no_high, df_no_high, "Suicide") |>
  mutate(sample = "Exclude >95th percentile")

curve_mh_main <- curve_mh |> mutate(sample = "Main sample")
curve_sui_main <- curve_sui |> mutate(sample = "Main sample")
```

```{r fig.width=8, fig.height=5}
ggplot(bind_rows(curve_mh_main, curve_mh_no_high), aes(pm25, RR, color = sample)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to extreme PM2.5 days: Mental health",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio"
  ) +
  theme_bw()
```

```{r fig.width=8, fig.height=5}
ggplot(bind_rows(curve_sui_main, curve_sui_no_high), aes(pm25, RR, color = sample)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to extreme PM2.5 days: Suicide",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio"
  ) +
  theme_bw()
```

## Sensitivity analysis 3: alternative time-trend df (3--6)

```{r}
fit_time_sensitivity <- function(outcome, time_df, data) {
  form <- as.formula(
    paste0(outcome,
           " ~ ns(pm25, 3) + ns(time, ", time_df, ") + ns(tmean, 3) + is_weekend + precip")
  )

  glm(form, family = quasipoisson(link = "log"), data = data)
}

time_df_results_mh <- map_df(3:6, function(k) {
  mod <- fit_time_sensitivity("mental_health", k, df)
  curve <- make_er_curve(mod, df, "Mental health")
  curve$time_df <- paste0("time df = ", k)
  curve
})

time_df_results_sui <- map_df(3:6, function(k) {
  mod <- fit_time_sensitivity("suicide", k, df)
  curve <- make_er_curve(mod, df, "Suicide")
  curve$time_df <- paste0("time df = ", k)
  curve
})
```

```{r fig.width=8, fig.height=5}
ggplot(time_df_results_mh, aes(pm25, RR, color = time_df)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to time-trend df: Mental health",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio",
    color = "Time spline"
  ) +
  theme_bw()
```

```{r fig.width=8, fig.height=5}
ggplot(time_df_results_sui, aes(pm25, RR, color = time_df)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to time-trend df: Suicide",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio",
    color = "Time spline"
  ) +
  theme_bw()
```

## Sensitivity analysis 4: nonlinear precipitation

```{r}
fit_mh_nl_precip <- glm(
  mental_health ~ ns(pm25, 3) + ns(time, 4) + ns(tmean, 3) + is_weekend + ns(precip, 3),
  family = quasipoisson(link = "log"),
  data = df
)

fit_sui_nl_precip <- glm(
  suicide ~ ns(pm25, 3) + ns(time, 4) + ns(tmean, 3) + is_weekend + ns(precip, 3),
  family = quasipoisson(link = "log"),
  data = df
)

curve_mh_nl_precip <- make_er_curve(fit_mh_nl_precip, df, "Mental health") |>
  mutate(model = "Nonlinear precip")

curve_sui_nl_precip <- make_er_curve(fit_sui_nl_precip, df, "Suicide") |>
  mutate(model = "Nonlinear precip")

curve_mh_main2 <- curve_mh |> mutate(model = "Linear precip")
curve_sui_main2 <- curve_sui |> mutate(model = "Linear precip")
```

```{r fig.width=8, fig.height=5}
ggplot(bind_rows(curve_mh_main2, curve_mh_nl_precip), aes(pm25, RR, color = model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to precipitation functional form: Mental health",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio"
  ) +
  theme_bw()
```

```{r fig.width=8, fig.height=5}
ggplot(bind_rows(curve_sui_main2, curve_sui_nl_precip), aes(pm25, RR, color = model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = 2) +
  labs(
    title = "Sensitivity to precipitation functional form: Suicide",
    x = "PM2.5 (\u03bcg/m^3)",
    y = "Rate ratio"
  ) +
  theme_bw()
```

## Summary table

```{r}
results_summary <- bind_rows(
  rr_mh_linear  |> mutate(outcome = "mental_health"),
  rr_sui_linear |> mutate(outcome = "suicide")
) |>
  select(outcome, RR_10, lower_95, upper_95)

results_summary
```


# Quasi-Poisson Time Series GLM (Yiming Cao)

## Model fitting

Using the same data loaded above. The main model keeps PM2.5 as a linear exposure term — the goal here is to estimate the rate ratio per 10 μg/m³ increase, adjusting for seasonal trend, temperature, weekend effect, and precipitation.

```{r}
library(splines)

# time variable as days since Jan 1
df$time_days <- as.numeric(df$date - min(df$date))

# mental health model
fit_qp_mh <- glm(mental_health ~ pm25 + ns(time_days, 4) + ns(tmean, 3) +
                    is_weekend + precip,
                  family = quasipoisson(link = "log"),
                  data = df)

# suicide model
fit_qp_sui <- glm(suicide ~ pm25 + ns(time_days, 4) + ns(tmean, 3) +
                     is_weekend + precip,
                   family = quasipoisson(link = "log"),
                   data = df)
```

```{r}
summary(fit_qp_mh)
```

```{r}
summary(fit_qp_sui)
```

## Dispersion check

```{r}
cat("Mental health dispersion:", round(summary(fit_qp_mh)$dispersion, 2), "\n")
cat("Suicide dispersion:", round(summary(fit_qp_sui)$dispersion, 2), "\n")
```

## Rate ratios per 10 μg/m³ PM2.5

```{r}
extract_rr <- function(model, label) {
  b <- coef(model)["pm25"]
  se <- sqrt(vcov(model)["pm25", "pm25"])
  data.frame(
    outcome = label,
    RR = exp(b * 10),
    lower = exp((b - 1.96 * se) * 10),
    upper = exp((b + 1.96 * se) * 10),
    p = summary(model)$coefficients["pm25", 4]
  )
}

rr_table <- rbind(
  extract_rr(fit_qp_mh, "Mental health"),
  extract_rr(fit_qp_sui, "Suicide")
)
rr_table$RR <- round(rr_table$RR, 3)
rr_table$lower <- round(rr_table$lower, 3)
rr_table$upper <- round(rr_table$upper, 3)
rr_table$p <- round(rr_table$p, 3)
rr_table
```

## Rate ratio forest plot

```{r fig.width=7, fig.height=3.5}
ggplot(rr_table, aes(x = RR, y = outcome)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.15) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  labs(x = "Rate ratio per 10 μg/m³ PM2.5", y = NULL,
       title = "PM2.5 association with EMS outcomes (Quasi-Poisson GLM)") +
  theme_bw(base_size = 11)
```

## Sensitivity analysis: time trend df (3–6)

```{r}
sens_df <- data.frame()
for (d in 3:6) {
  fit_tmp <- glm(mental_health ~ pm25 + ns(time_days, d) + ns(tmean, 3) +
                   is_weekend + precip,
                 family = quasipoisson, data = df)
  b <- coef(fit_tmp)["pm25"]
  se <- sqrt(vcov(fit_tmp)["pm25", "pm25"])

  sens_df <- rbind(sens_df, data.frame(
    df_time = d,
    outcome = "Mental health",
    RR = exp(b * 10),
    lower = exp((b - 1.96 * se) * 10),
    upper = exp((b + 1.96 * se) * 10)
  ))

  fit_tmp2 <- glm(suicide ~ pm25 + ns(time_days, d) + ns(tmean, 3) +
                     is_weekend + precip,
                   family = quasipoisson, data = df)
  b2 <- coef(fit_tmp2)["pm25"]
  se2 <- sqrt(vcov(fit_tmp2)["pm25", "pm25"])

  sens_df <- rbind(sens_df, data.frame(
    df_time = d,
    outcome = "Suicide",
    RR = exp(b2 * 10),
    lower = exp((b2 - 1.96 * se2) * 10),
    upper = exp((b2 + 1.96 * se2) * 10)
  ))
}
```

```{r fig.width=8, fig.height=4.5}
ggplot(sens_df, aes(x = factor(df_time), y = RR)) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  facet_wrap(~outcome, scales = "free_y") +
  labs(x = "Time trend spline df", y = "Rate ratio per 10 μg/m³ PM2.5",
       title = "Sensitivity of PM2.5 effect to time trend flexibility") +
  theme_bw(base_size = 11)
```

## Sensitivity analysis: exclude high PM2.5 days

```{r}
pm95_cutoff <- quantile(df$pm25, 0.95)
df_trimmed <- df[df$pm25 <= pm95_cutoff, ]
cat("Excluded", nrow(df) - nrow(df_trimmed), "days with PM2.5 >",
    round(pm95_cutoff, 1), "μg/m³\n")

fit_trim_mh <- glm(mental_health ~ pm25 + ns(time_days, 4) + ns(tmean, 3) +
                      is_weekend + precip,
                    family = quasipoisson, data = df_trimmed)
fit_trim_sui <- glm(suicide ~ pm25 + ns(time_days, 4) + ns(tmean, 3) +
                       is_weekend + precip,
                     family = quasipoisson, data = df_trimmed)

rr_trim <- rbind(
  extract_rr(fit_trim_mh, "Mental health (trimmed)"),
  extract_rr(fit_trim_sui, "Suicide (trimmed)")
)
rr_trim$RR <- round(rr_trim$RR, 3)
rr_trim$lower <- round(rr_trim$lower, 3)
rr_trim$upper <- round(rr_trim$upper, 3)
rr_trim$p <- round(rr_trim$p, 3)

rbind(rr_table, rr_trim)
```

## Residual diagnostics

```{r fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
plot(fitted(fit_qp_mh), residuals(fit_qp_mh, type = "deviance"),
     xlab = "Fitted", ylab = "Deviance residuals",
     main = "Mental health: residuals vs fitted", pch = 16, cex = 0.7)
abline(h = 0, lty = 2)

plot(df$date, residuals(fit_qp_mh, type = "deviance"),
     xlab = "Date", ylab = "Deviance residuals",
     main = "Mental health: residuals over time", pch = 16, cex = 0.7)
abline(h = 0, lty = 2)
```

```{r fig.width=7, fig.height=4}
pacf(residuals(fit_qp_mh, type = "deviance"),
     main = "PACF of deviance residuals (mental health)")
```

```{r fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
plot(fitted(fit_qp_sui), residuals(fit_qp_sui, type = "deviance"),
     xlab = "Fitted", ylab = "Deviance residuals",
     main = "Suicide: residuals vs fitted", pch = 16, cex = 0.7)
abline(h = 0, lty = 2)

plot(df$date, residuals(fit_qp_sui, type = "deviance"),
     xlab = "Date", ylab = "Deviance residuals",
     main = "Suicide: residuals over time", pch = 16, cex = 0.7)
abline(h = 0, lty = 2)
```

## Predicted vs observed

```{r fig.width=9, fig.height=4}
df$pred_mh <- fitted(fit_qp_mh)

ggplot(df, aes(x = date)) +
  geom_point(aes(y = mental_health), alpha = 0.4, size = 1.2, color = "gray40") +
  geom_line(aes(y = pred_mh), color = "#E15759", linewidth = 0.8) +
  labs(x = "Date", y = "Daily count",
       title = "Mental health calls: observed vs quasi-Poisson fitted values") +
  theme_bw(base_size = 11)
```

## Results section outline (Quasi-Poisson GLM)

The quasi-Poisson GLM estimated a rate ratio of 0.974 (95% CI: 0.932–1.017) per 10 μg/m³ increase in same-day PM2.5 for mental health EMS calls, after adjusting for seasonal trend, temperature, weekend, and precipitation (p = 0.228). For suicide-related calls, the estimated rate ratio was 1.089 (95% CI: 0.837–1.416, p = 0.526). Neither association was statistically significant.

Key points for the results write-up:

- The dispersion parameter for mental health was 2.24, confirming overdispersion relative to standard Poisson. For suicide calls the dispersion was 0.95, close to 1.
- The PM2.5 coefficient was not statistically significant for either outcome.
- Weekend indicator was a strong predictor for mental health calls (lower on weekends, p < 0.001), consistent with the EDA.
- Sensitivity analysis: the PM2.5 rate ratio for mental health was stable across time trend df = 3–6 (ranging from 0.972 to 0.976). However, excluding high-PM2.5 days (>95th percentile) shifted the estimate from 0.974 to 1.015, suggesting the slight negative association in the main model may be driven by a few high-exposure days.
- Residual diagnostics: check PACF plot for any remaining temporal autocorrelation after the time spline adjustment.

